# Descriptive Statistics

## Introduction

In this lesson, we'll delve into the world of descriptive statistics. Descriptive statistics provide a summary of your data, allowing you to understand its central tendency, variability, and shape. We'll learn how to calculate common descriptive statistics in R and interpret their meaning.

## Learning Objectives

*   Calculate measures of central tendency (mean, median, mode).
*   Calculate measures of dispersion (standard deviation, variance, IQR, range).
*   Use `dplyr::summarize()` to efficiently calculate descriptive statistics.
*   Understand the relationship between descriptive statistics and data distribution.
*   Calculate descriptive statistics for your cloud hosted R project.
*   Commit changes to cloud hosted R project with descriptive statistics.

## Measures of Central Tendency

Measures of central tendency describe the "center" of a dataset.

1.  **Mean:** The average of all values.
    *   Calculated as the sum of the values divided by the number of values.

    ```{r}
    scores <- c(75, 80, 92, 68, 85)
    mean(scores)
    ```

2.  **Median:** The middle value when the data is sorted.
    *   If there are an even number of values, the median is the average of the two middle values.

    ```{r}
    median(scores)
    ```

3.  **Mode:** The most frequent value in the dataset.
    *   R doesn't have a built-in function to calculate the mode directly, so we'll create one:

    ```{r}
    getmode <- function(v) {
      uniqv <- unique(v)
      uniqv[which.max(tabulate(match(v, uniqv)))]
    }

    values <- c(1, 2, 2, 3, 4, 4, 4, 5)
    getmode(values)
    ```

**When to Use Which Measure:**

*   **Mean:** Sensitive to outliers (extreme values). Use when the data is relatively symmetrical and has no extreme outliers.
*   **Median:** Robust to outliers. Use when the data has outliers or is skewed.
*   **Mode:** Useful for categorical data or discrete data with repeating values.

## Measures of Dispersion

Measures of dispersion describe the spread or variability of a dataset.

1.  **Standard Deviation:** A measure of how spread out the data is around the mean.
    *   A higher standard deviation indicates greater variability.

    ```{r}
    sd(scores)
    ```

2.  **Variance:** The square of the standard deviation.
    *   Also measures variability, but is less interpretable than standard deviation.

    ```{r}
    var(scores)
    ```

3.  **Interquartile Range (IQR):** The difference between the 75th percentile (Q3) and the 25th percentile (Q1).
    *   Robust to outliers and provides a measure of the spread of the middle 50% of the data.

    ```{r}
    IQR(scores)
    ```

4.  **Range:** The difference between the maximum and minimum values.

    ```{r}
    range(scores)
    diff(range(scores)) #Calculate the range from the output
    ```

## Descriptive Statistics with `dplyr::summarize()`

The `dplyr` package provides the `summarize()` function, which makes it easy to calculate multiple descriptive statistics at once:

```{r}
library(dplyr)

#Replace this file with your local file.
exam_scores <- read.csv("https://raw.githubusercontent.com/sijuswamyresearch/R-for-Data-Analytics/refs/heads/main/data/exam_scores.csv")

exam_scores %>%
  summarize(
    mean_score = mean(score),
    median_score = median(score),
    sd_score = sd(score),
    iqr_score = IQR(score),
    min_score = min(score),
    max_score = max(score)
  )
```

### Grouped Descriptive Statistics

You can calculate descriptive statistics for different groups within your data using `dplyr::group_by()` in combination with `summarize()`:

```{r}
exam_scores %>%
  group_by(grade) %>%
  summarize(
    mean_score = mean(score),
    median_score = median(score),
    sd_score = sd(score)
  )
```

## Descriptive Statistics and Data Distribution

Descriptive statistics provide insights into the distribution of your data:

*   **Symmetrical Distribution:**
    *   Mean ≈ Median ≈ Mode
    *   Standard deviation is relatively small.
*   **Right-Skewed Distribution:**
    *   Mean > Median > Mode
    *   Long tail on the right side.
*   **Left-Skewed Distribution:**
    *   Mean < Median < Mode
    *   Long tail on the left side.

## Practice

1.  Load the `exam_scores.csv` dataset (or your own dataset).
2.  Calculate the mean, median, standard deviation, IQR, and range for a numerical column.
3.  Create a custom function to calculate the mode.
4.  Calculate descriptive statistics for different groups within the data (e.g., by gender, by region).
5.  Describe the shape of the data based on the descriptive statistics you calculated.